---
title: "Spectral reflectance curves of natural habitats of Southern Bug river valley, Ukraine"
author: "Oleh Prylutskyi"
date: "9/05/2021"
output: html_document
---
Reset R`s brain

```{r}
rm(list = ls())
```

# Load required packages and data
Load packages

```{r}
library(plyr)# order matters! Firstly plyr, followed by dplyr!
library(dplyr)
library(tidyr)
library(raster)
library(sf)
library(sp)
library(rgdal)
library(exactextractr)
library(ggplot2)
library(mgcv) # for geom_smooth by method = gam
library(Hmisc) # for stat_summary
```

We use several time spans which represent different phenological stages, as following
[Day of the year calendar](https://people.biology.ucsd.edu/patrick/julian_cal.html)
Be careful with leap years!

## Vegetative periods
measured in day of year

105-135 - (15 Apr - 15 May) 
135-166 - (15 May - 15 Jun)
166-196 - (15 Jun - 15 Jul)
196-227 - (15 Jul - 15 Aug)
227-258 - (15 Aug - 15 Sep)
258-288 - (15 Sep - 15 Oct)

## leaf-less preriods

288-319 - (15 Oct - 15 Nov)
074-105 - (15 Mar - 15 Apr)

Set timespan names. For each timespan the code should be run separately.

```{r}
year = 2019
timespan = "_15May-15Jun"
time_span = "135_166"
```

Load shapefile with ground truth data (as polygons with known habitat type) and check number of polygons for each habitat types

```{r}
shp <- shapefile("training_polygons_shapefile.shp")
summary(factor(shp@data$eunis_2020)) # number of training polygons
```

```{r}
Classes = cbind(levels(factor(shp@data$eunis_2020)),c(1:26))
colnames(Classes) = c("eunis_2020", "ClassInt")
Classes = as.data.frame(Classes)
Classes$ClassInt = as.numeric(Classes$ClassInt)
shp = merge(shp, Classes, by.x = "eunis_2020")
```

Set the satellite imagery name

```{r}
imageryname = paste0("SBuh_S2_full_",year, "_", time_span)
```
Define imagery to process
Load the raster file

```{r}
filename = paste0(imageryname,".tif")
ImageStack <- brick(filename)
ImageStack # View imagery structure and features
```

Extract optical data from satellite imagery for each pixels within training polygons and convert it to the dataframe

```{r}
dfAll <- data.frame(matrix(vector(), nrow = 0, ncol = length(names(ImageStack)) + 1))
for (i in 1:length(base::unique(shp[["ClassInt"]]))){
  category <- base::unique(shp[["ClassInt"]])[i]
  categorymap <- shp[shp[["ClassInt"]] == category,]
  beginCluster()
  dataSet <- exact_extract(ImageStack, categorymap)
  endCluster()
  dataSet <- dataSet[!unlist(lapply(dataSet, is.null))]
  dataSet <- lapply(dataSet, function(x){cbind(x, ClassInt = as.numeric(rep(category, nrow(x))))})
  df <- do.call("rbind", dataSet)
  dfAll <- rbind(dfAll, df)
}
```

Add classes' names to resulting dataframe and it for further sessions.

```{r}
reflectance = dfAll %>%
  left_join(Classes)
save(reflectance, file = imageryname)
```

# Make spectral reflectance curve for single imagery

## Quantitative overview of pixel data

```{r}
summary(factor(reflectance$eunis_2020))
```

Number of spectral values for different habitat types

```{r}
ggplot(reflectance, aes(x=eunis_2020))+
  geom_bar()+
  theme_minimal()
```

Define which habitat types will be excluded, such as small habitats lacking of pixel data.

```{r}
excluded = c("V34", "Q53", "S91") # manualy define habitats with less than 50 values
```

## Prepare pixel data to the ready-to-visualize mode

Create "dummy" wavelength object, containing mean wavelengths (nm) for Sentinel 2A (https://en.wikipedia.org/wiki/Sentinel-2), for bands 2-12

```{r}
dummy_wavelength = c(492.4, 559.8, 664.6, 704.1, 740.5, 782.8, 832.8, 864.7, 1613.7, 2202.4)
bands = c("B2", "B3", "B4", "B5", "B6", "B7", "B8", "B8A", "B11", "B12")
waves = cbind(bands, dummy_wavelength)
colnames(waves)[1] = "variable"
```

For Sentinel-2 L2A optical data, generated through Google Earth Engine, the relation between DN and REFLECTANCE (default unit) is: DN = 10000 * REFLECTANCE. [documentation](https://docs.sentinel-hub.com/api/latest/data/sentinel-2-l2a/). So, to obtain unified reflectance data, we need to divide pixel values to 10,000.

Separate EUNIS habitat types of high hierarchical level

```{r}
EUNIS = substring(Classes$eunis_2020, 1, 1) # split first character from strings
Classes_2l = cbind(Classes, EUNIS)
```

Obtaining true reflectance values and add

```{r}
df = reflectance %>% dplyr::select(-eunis_2020, -NDVI) %>%
  mutate_at(vars(B2: B12), funs(./ 10000)) %>% # obtain true reflectance values
  left_join(Classes_2l) %>% # Add EUNIS classification systems on 2 levels
  filter(!(eunis_2020 %in% excluded)) # drop enormous habitat types
```

Quantitative overview of resulting pixel data for both hierarchical level

```{r For higher hierarchical level}
summary(factor(df$EUNIS))
```

```{r For lower hierarchical level}
summary(factor(df$eunis_2020))
```

Reshape the dataframe to make it appropriate to ggplot2 syntax

```{r}
df1 = tibble::as_tibble(df) %>%
  reshape2::melt(id = "eunis_2020") %>%
  left_join(as.data.frame(waves)) %>%
  left_join(Classes_2l) %>%
  mutate(across(eunis_2020, as.factor)) %>%
  mutate(across(dummy_wavelength, as.numeric)) %>%
  mutate(across(variable, as.factor)) %>%
  mutate(across(value, as.numeric)) %>%
  mutate(across(EUNIS, as.factor)) %>%
  mutate(variable = factor(variable, 
                       levels = c("B2","B3","B4","B5","B6","B7","B8","B8A","B11","B12"))) %>%
  na.omit()
```

Spectral reflectance curves for all habitat types

```{r}
filnm = paste0("SRC_all_habitats_",as.character(year),timespan,".png")
png(filnm, width = 200,
      height = 170, units = "mm", res = 150)
ggplot(df1, aes(x=dummy_wavelength, y= value, colour = eunis_2020))+
  geom_smooth(aes(fill = EUNIS))+
  ylim(0, 0.5)+
  labs(x='Wavelength, nm', y='Reflectance',
       title=paste0("Spectral reflectance curves for different habitat types ",year,timespan),
       caption='Data: Sentinel-2 Level-2A')+
  theme_minimal()
dev.off()
```

Within different types on higher hierarchical level (R, J, etc.)

```{r}
filnm = paste0("SRC_all_habitats_facet_",as.character(year),timespan,".png")
png(filnm, width = 200,
      height = 150, units = "mm", res = 150)
ggplot(df1, aes(x=dummy_wavelength, y= value, colour = eunis_2020))+
  geom_smooth()+
  ylim(0, 0.5)+
  labs(x='Wavelength, nm', y='Reflectance',
       title=paste0("Spectral reflectance curves for different habitat types ",year,timespan),
       caption='Data: Sentinel-2 Level-2A')+
  facet_wrap(~ EUNIS)+
  theme_minimal()
dev.off()
```

Transform the data to acquire statistical summaries (mean, mean-standard deviation, mean+standard deviation) for each group of variables

```{r}
# Summarize
df3 = df1 %>%
  drop_na() %>%
  dplyr::select(-ClassInt) %>%
  mutate(across(variable, as.factor)) %>%
  mutate(across(value, as.numeric)) %>%
  mutate(across(EUNIS, as.factor)) %>%
  group_by(variable, EUNIS, eunis_2020) %>% 
  summarise(mean_refl = mean(value), min_refl = mean(value)-sd(value), max_refl = mean(value)+sd(value)) %>%
  left_join(as.data.frame(waves)) %>%
  mutate(across(dummy_wavelength, as.numeric)) %>%
  rename(band = variable, wavelength = dummy_wavelength) %>%
  mutate(band = factor(band, 
                       levels = c("B2","B3","B4","B5","B6","B7","B8","B8A","B11","B12")))
```


Reflectance ber bands on high classification level

```{r}
filnm = paste0("Reflectance_per_bands_EUNIS_",as.character(year),timespan,".png")
png(filnm, width = 200,
      height = 150, units = "mm", res = 150)
ggplot(df3, aes(x=band, y=mean_refl, colour = EUNIS))+
  geom_line(aes(group = eunis_2020))+
  geom_pointrange(aes(ymin = min_refl, ymax = max_refl), width = 0.2)+
  ylim(0, 0.6)+
  labs(x='Sentinel-2 bands', y='Reflectance',
       title=paste0("Reflectance for different habitat types ",year,timespan),
       caption='Data: Sentinel-2 Level-2A')+
  theme_minimal()
dev.off()
```

Reflectance per bands within different groups of habitat types

```{r}
filnm = paste0("Reflectance_per_bands_EUNIS_facet_",as.character(year),timespan,".png")
png(filnm, width = 200,
      height = 150, units = "mm", res = 150)
ggplot(df3, aes(x=band, y=mean_refl, colour = eunis_2020))+
  geom_line(aes(group = eunis_2020))+
  geom_pointrange(aes(ymin = min_refl, ymax = max_refl), width = 0.2)+
  ylim(0, 0.6)+
  labs(x='Sentinel-2 bands', y='Reflectance',
       title=paste0("Reflectance for different habitat types ",year,timespan),
       caption='Data: Sentinel-2 Level-2A')+
  facet_wrap(~ EUNIS)+
  theme_minimal()
dev.off()
```

# Differences between habitat types per channel
For 1st hierarchical level (EUNIS)

```{r}
png("Violin_all_variables_EUNIS.png", width = 250,
      height = 210, units = "mm", res = 150)
ggplot(df1, aes(x=EUNIS, y= value, fill = EUNIS))+
  geom_violin(trim = FALSE)+
  facet_wrap( ~ variable, ncol = 2)+
  ylim(0, 0.6)+
  labs(x='Habitat type',y='Reflectance',
       fill="EUNIS",
       title=paste0("Reflectance for different habitat types ",year,timespan),
       caption='Data: Sentinel-2 Level-2A')+
  theme_minimal()
dev.off()
```

For lower hierarchical level (eunis_2020)

```{r}
png("Violin_all_variables_eunis_2020.png", width = 300,
      height = 450, units = "mm", res = 150)
ggplot(df1, aes(x=eunis_2020, y= value, fill = eunis_2020))+
  geom_violin(trim = FALSE)+
  facet_wrap( ~ variable, ncol = 1)+
  ylim(0., 0.6)+
  labs(x='Habitat type',y='Reflectance',
       fill="eunis_2020",
       title=paste0("Reflectance for different habitat types ",year,timespan),
       caption='Data: Sentinel-2 Level-2A')+
  theme_minimal()
dev.off()
```

# Reflectance within particular hirher level (R, S, etc.)

```{r}
#habtype = "C" # Uncomment desired, leave commented others
#habtype = "J"
#habtype = "Q"
habtype = "R"
#habtype = "S"
#habtype = "T"
#habtype = "U"
#habtype = "V"
#habtype = "X"

# As violin plots
filnm = paste0("Violins_",habtype,"_",as.character(year),timespan,".png")
png(filnm, width = 200,
      height = 150, units = "mm", res = 150)
df1 %>%
  filter(EUNIS == habtype) %>%
  ggplot(aes(x=eunis_2020, y= value, fill = eunis_2020))+
  geom_violin(trim = FALSE)+
  facet_wrap( ~ variable, ncol = 3)+
  labs(x='Habitat type',y='Reflectance',
       fill="eunis_2020",
       title=paste0("Reflectance for different habitat types ",year,timespan),
       caption='Data: Sentinel-2 Level-2A')+
  theme_minimal()
dev.off()

# As Spectral reflectance curves
filnm = paste0("SRC_",habtype,"_",as.character(year),timespan,".png")
png(filnm, width = 160,
      height = 120, units = "mm", res = 150)
df1 %>%
  filter(EUNIS == habtype) %>%
  ggplot(aes(x=dummy_wavelength, y= value, colour = eunis_2020))+
  geom_smooth()+
  ylim(0, 0.5)+
  labs(x='Wavelength, nm', y='Reflectance',
       title=paste0("Reflectance for different habitat types ",year,timespan),
       caption='Data: Sentinel-2 Level-2A')+
  theme_minimal()
dev.off()

# As reflectance per bands
filnm = paste0("Reflectance_per_bands_",habtype,"_",as.character(year),timespan,".png")
png(filnm, width = 160,
      height = 120, units = "mm", res = 150)
df3 %>%
  filter(EUNIS == habtype) %>%
  ggplot(aes(x=band, y=mean_refl, colour = eunis_2020))+
  geom_line(aes(group = eunis_2020))+
  ylim(0,0.6)+
  geom_pointrange(aes(ymin = min_refl, ymax = max_refl, alpha = 0.2), width = 0.2)+
  labs(x='Sentinel-2 bands', y='Reflectance',
       title=paste0("Reflectance for different habitat types ",year,timespan),
       caption='Data: Sentinel-2 Level-2A')+
  theme_minimal()
dev.off()
```

# Comparison for some closelly related groups of habitat types
## Wooded speppe - tue steppe

- R1B - Steppe (dry grassland)
- S36 - Low steppic scrub
- S35 - Temperate thorn scrub
- X18 - Wooded steppe
- T19 - Termophilous forest
- Q51 - Tall-helophyte bed
- R21 - Mesic lowland pasture
- T11 - Riparian forest
- R12 - Cryptogam- and annual-dominated vegetation on siliceous rock outcrops
- U33 - Temperate siliceous inland cliff

Set habitat_set variable (above) - uncomment desired or add yours

```{r}
habitat_set = c("S35","S36","R1B","X18","T19")
#habitat_set = c("Q51","T11","R21")
#habitat_set = c("U33","R12","R1B")

habset = paste(habitat_set, collapse = "_") # convert vector to string

# Statistical summary for reflectance for selected habitat types
filnm = paste0("Reflectance_within_bands_",habset,"_",as.character(year),timespan,".png")
png(filnm, width = 160,
      height = 120, units = "mm", res = 150)
df3 %>%
  filter(eunis_2020 %in% habitat_set) %>% # set criteria for subsetting
  ggplot(aes(x=band, y=mean_refl, colour = eunis_2020))+
  geom_line(aes(group = eunis_2020))+
  geom_pointrange(aes(ymin = min_refl, ymax = max_refl, alpha = 0.2), width = 0.2)+
  ylim(0, 0.6)+
  labs(x='Sentinel-2 bands', y='Reflectance',
       title=paste0("Reflectance for different habitat types ",year,timespan),
       caption='Data: Sentinel-2 Level-2A')+
  theme_minimal()
dev.off()

# Spectral reflectance curves for selected habitat types
filnm = paste0("SRC_within_bands_",habset,"_",as.character(year),timespan,".png")
png(filnm, width = 160,
      height = 120, units = "mm", res = 150)
df1 %>%
  filter(eunis_2020 %in% habitat_set) %>% # set criteria for subsetting
  ggplot(aes(x=dummy_wavelength, y= value, colour = eunis_2020))+
  geom_smooth()+
  ylim(0, 0.5)+
  labs(x='Wavelength, nm', y='Reflectance',
       title=paste0("Reflectance for different habitat types ",year,timespan),
       caption='Data: Sentinel-2 Level-2A')+
  theme_minimal()
dev.off()

# Violin plots of reflectance value density for selected habitat types
filnm = paste0("Violins_within_bands_",habset,"_",as.character(year),timespan,".png")
png(filnm, width = 200,
      height = 120, units = "mm", res = 150)
df1 %>%
  filter(eunis_2020 %in% habitat_set) %>%
  ggplot(aes(x=variable, y=value, fill=eunis_2020))+
  geom_violin(trim = FALSE)+
  labs(x='Sentinel-2 bands', y='Reflectance',
       title=paste0("Reflectance for different habitat types ",year,timespan),
       caption='Data: Sentinel-2 Level-2A')+
  theme_minimal()
dev.off()
```

# Save transformed reflectance data for further analysis

```{r}
dataframename = "df1"
dataname = paste0(dataframename,"_",imageryname)
save(df1, file = dataname)
dataframename = "df3"
dataname = paste0(dataframename,"_",imageryname)
save(df3, file = dataname)
```

# Obtain NDVI values

```{r}
df = reflectance %>% dplyr::select(eunis_2020, NDVI) %>%
  mutate(NDVI = NDVI/1000) %>% # obtain true reflectance values
  left_join(Classes_2l) %>% # Add EUNIS classification systems on 2 levels
  filter(!(eunis_2020 %in% excluded)) %>% # drop enormous habitat types
  dplyr::select(-ClassInt) %>%
  mutate(Year = year) %>%
  mutate(across(Year, as.factor)) %>%
  mutate(across(EUNIS, as.factor)) %>%
  mutate(across(eunis_2020, as.factor))
save(df, file = paste0("NDVI_",year,"_",time_span))
```
